---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### loading the data
Here we decide to load the dataset directly from the source url where the file is store in the web. This strategy ensure that there is no manual manipulation of the data source. 
```{r loading, echo = TRUE} 
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip",temp)
data <- read.csv(unz(temp, "activity.csv"))
unlink(temp)
```

### Preprocessing the data
Before begining the analysis we need to  Convert the date variable from character format to Date format. As we will use  mutate() verb from **dplyr**, we need first to load the package.

```{r, message=FALSE} 
library(dplyr)
```

Converting date from character to date format. To do this we save the data in a new data frame **mydf**
```{r processing, echo = TRUE} 
mydf <- mutate(data, date = as.Date(strptime(date, "%Y-%m-%d"))) 
```

## What is mean total number of steps taken per day?

1. Number of steps taken per day.

We generate a data frame ** stepbydate ** with to variables *steps* and *date*. each observations consist of a date and the corresponding total number of steps.
```{r, echo = TRUE}
stepbydate <- mydf %>%group_by(date) %>% summarize(steps=sum(steps,na.rm=TRUE)) 
```

2. Make a histogram of the total number of steps taken each day. 
```{r histogram1, echo = TRUE}
 hist(stepbydate$steps, main = "Total number of steps taken each day", xlab='Number of steps', ylab="Frequency",col="green", breaks = 20) 
```
 

3. Mean and median of the total number of steps taken per day
```{r mean_median1, echo = TRUE}
summarize(stepbydate, mean_steps=mean(steps,na.rm=TRUE), median_steps=median(steps,na.rm=TRUE))
```

## What is the average daily activity pattern?

1. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
To do this plot, we first generate a new data frame **mydf2**
```{r, echo = TRUE}
 mydf2 <- mydf %>% group_by(interval) %>% summarize(steps=mean(steps,na.rm=TRUE)) 
```

The we plot resulting time serie
```{r plot1, echo = TRUE}
 plot(mydf2$interval, mydf2$steps, type = "l", lwd=2, col="green", xlab="interval", ylab="Average steps")
```


2. Interval on average across all the days in the dataset which contains the maximum number of steps
To answer to this question we use the select function from **sqldf** package. First we load the package
```{r, message=FALSE}
library(sqldf)
```

We can then use **select()** function to find the maximum number of steps
```{r, echo = TRUE}
sqldf("select interval, max(steps) steps from mydf2")
```

## Imputing missing values

1. Total number of missing values in the dataset 
```{r, echo = TRUE}
 nbna <- sum(!complete.cases(data))
```

2. Strategy for filling in all of the missing values in the dataset

Here we answer at the same time to the  questions 2 and 3. 

Here is our strategy: We choose to fill na's with the mean for that 5-minute interval. For each 5-minute interval, we compute the mean without missing values in all the dataset. The value obtained is use to fill missing values in the same 5-minute interval. 
```{r strategy, echo = TRUE}
mydf3 <- mydf %>% group_by(interval) %>% mutate(filled_steps=mean(steps,na.rm=TRUE)) %>% transform(steps = ifelse(is.na(steps), filled_steps, steps)) %>% select(steps,date,interval)
```

4.  Make a histogram of the total number of steps taken each day
We generate a data frame ** stepbydate ** with to variables *steps* and *date*. each observations consist of a date and the corresponding total number of steps.
```{r, echo = TRUE}
 stepbydate <- mydf3 %>%group_by(date) %>% summarize(steps=sum(steps,na.rm=TRUE)) 
```

Making the histogram of the total number of steps taken each day. 
```{r histogram2, echo = TRUE}
 hist(stepbydate$steps, main = "Total number of steps taken each day", xlab='Number of steps', ylab="Frequency",col="green", breaks = 20)
```

 Mean and median of the total number of steps taken per day
```{r mean_median2, echo = TRUE} 
summarize(stepbydate, mean_steps=mean(steps,na.rm=TRUE), median_steps=median(steps,na.rm=TRUE))
``` 
 Now the mean and the median are the same : **10766.19** 
 These values differ from the estimates from the first part of the assignment wich were **9354.23 for the mean** and **10395 for the median**
 The median and the mean are moving to the right

## Are there differences in activity patterns between weekdays and weekends?
1. Creating a new factor variable in the dataset consisting of two factors: *"weekend","weekday"*
```{r, echo = TRUE}
 mydf4 <- mydf3 %>% mutate(dayname=weekdays(date)) %>% mutate(isweekday=ifelse(dayname=="samedi"|dayname=="dimanche","weekend","weekday")) %>% select(steps,date,interval,isweekday)
```

2. Panel plot of time series plot of the 5-minute interval  and the average number of steps
We first generate a new data frame **mydf5** with three variables: *steps* (mean of steps by interval), *interval* and *isweekday*
```{r, echo = TRUE}
  mydf5 <- mydf4 %>% group_by(interval,isweekday) %>% summarize(steps=mean(steps)) %>% select(steps,interval,isweekday)
```

As we will use **ggplot2** for this plot, we first need to load the package.
```{r, echo = TRUE}
 library(ggplot2)
```

Finally, we can make the plot
```{r plot2, echo = TRUE}
 g <- ggplot(mydf4, aes(x = interval, y = steps, group = isweekday)) + geom_line() + facet_grid(isweekday ~ .)
 g
```

We find that there are differences in activity patterns. During weekdays, on average the number of steps are above 600. This is different from the weekend where just a few observations have  average number of steps above 600.
